---
title: 'SECDB - Database Initialization Notebook'
author: "H. David Shea"
date: 2021-02-09
output:
    github_document: default
---

```{r setup, include = FALSE}
library(DBI)
library(RSQLite)
library(tidyverse)

base_dir <- here::here("")
db_file <- fs::path(base_dir, "SECDB")

if(dbCanConnect(RSQLite::SQLite(), db_file)) {
    secdb <- dbConnect(RSQLite::SQLite(), db_file)
}

knitr::opts_chunk$set(
    connection = "secdb", # automatically uses this connection in sql chunks 
    comment = "#>", 
    collapse = TRUE, 
    message = FALSE,
    fig.width = 8,
    fig.asp = ((1 + sqrt(5)) / 2) - 1, # the golden ratio - technically, the b proportion of a+b when a is 1
    out.width = "70%",
    fig.align = "center"
)
```

## Introduction

This notebook documents **and** initializes the `SECDB` basic stock research database. Only the definition and group constituent tables are initialized with the chunks processed in this notebook. However, all tables are dropped and recreated when running the processing chunks. The database is currently designed to store stock research data for a large section of US listed and traded stocks - roughly representing the stocks in the S&P 1500 Index (SP1500). This initialization procedure assumes that the initiation date (`start_date` in time dependent definition and group constituent tables) for all data in the SECDB database is *today*.  

NOTE:  This will be adjusted when we - later - load the pricing data for securities and get a better feel for the actual time frame for securities.

Sidebar: This notebook will also serve to document/experiment with mixing multiple execution engines within a single R Markdown document. Specifically, I will use the `bash`, `sql`, and `r` engines in various processing chunks within this notebook. One of the drawbacks that I have seen so far with chunks using the `sql` engine, is that only single `SQL` statements can be executed within the chunk. So, when many `SQL` statements are *logically* required together - for instance when dropping or recreating the entire schema - I have included the `SQL` statements in a script file and process them via the `SQLite` command within a `bash` engine chunk.

## Recreate SECDB

The next two code chunks process `SQL` scripts through the `sqlite3 SECDB` command via the `bash` engine.

This chunk references `DROP_SECDB_SCHEMA.SQL` as input to drop all of the schema objects for `SECDB`. Note that if the schema objects do not exist, you will see warning messages.

```{bash drop_schema, echo = TRUE}
sqlite3 SECDB < DROP_SECDB_SCHEMA.SQL
```

This chunk references `CREATE_SECDB_SCHEMA.SQL` as input to create all of the schema objects for `SECDB`.

```{bash create_schema, echo = TRUE}
sqlite3 SECDB < CREATE_SECDB_SCHEMA.SQL
```

The next chunk - which is an `r` chunk - uses the `RSQLite` function `dbListTables` to verify that the tables have been created.

```{r list_tables}
dbListTables(secdb)
```

## Initialize base SECDB definition and group constituent tables

As mentioned, the SECDB database is currently designed to store stock research data for - roughly - the stocks in the S&P 1500 Index (SP1500). 

The base data that we will populate within this notebook are populated with data sourced from wikipedia. The `stock_data_access.R` script within this project collects the various data required by scraping various web pages. These data include descriptive data for the securities in the SP500, SP400 and SP600 - collectively the SP1500 and industry classification data for the securities.

The script is sourced below so that the tibbles produced are available for loading into the appropriate base tables.

NOTE:  There is a portion of this code which accesses Yahoo Finance which takes a long time to process and - currently - has limited utility for sourcing sub-industry classifications for some older securities with missing classifications.  Looking for a better solution currently.

```{r stock_data_script}
source(fs::path(base_dir, "stock_data_access.R"))
```

The data in the gics table are Global Industry Classification Standard (GICS®) data developed by MSCI and Standard & Poor's in 1999. They establish a hierarchy of sector, industry group, industry and sub-industry classifications for a broad array of global stocks.

For the GICS data, the tibbles `sct_tbl`, `igp_tbl`, `ind_tbl` and `sub_tbl` contain data for GICS sectors, industry groups, industries and sub-industries, respectively. They all have the same column structure. The following `r` chunk uses the base `DBI` function `dbAppendTable` to `INSERT` the data from the four tibbles.

```{r load_gics_data}
dbAppendTable(secdb, "gics", sct_tbl)

dbAppendTable(secdb, "gics", igp_tbl)

dbAppendTable(secdb, "gics", ind_tbl)

dbAppendTable(secdb, "gics", sub_tbl)
```

Note that the `dbAppendTable` calls return the number of rows inserted into the table for each call. And, we can check with the following raw `sql` chunks to verify that the data are loaded correctly.

```{sql sct_select}
SELECT *
FROM   gics
WHERE  level = 'SCT';
```

```{sql igp_select}
SELECT * 
FROM   gics
WHERE  level = 'IGP';
```

```{sql ind_select}
SELECT * 
FROM   gics
WHERE  level = 'IND';
```

```{sql sub_select}
SELECT *
FROM   gics
WHERE  level = 'SUB';
```

Note that for `sql` chunks, the default in R Markdown is to display the first 10 records returned by `SELECT` statements in a document. This can be changed with the chunk option `max.print` which can be set equal to a specific number or to no limit by using `max.print = -1` or `max.print = NA`.

For descriptive security data, the above script pulled data for the S&P 1500 index constituents.

Data for the SP1500 are contained in the tibble `sp1500_tbl`. For the initialization, the sourced file above assumed that the `security` table was empty and created `uid` values explicitly. This is actually required because the auto-populate primary key functionality in `SQLite` only works when there is a single integer field primary key. The primary key for the `security` table references the `uid` and the `start_date` fields.

*[TODO: discuss strategy for future insertions]*

The data in the `sp500_tbl` mimic the tables structure for the `security` table which is required in order to use the `dbAppendTable` function.

```{r review_sp500_tbl}
sp1500_tbl
```

The following `r` chunk uses the base `DBI` function `dbAppendTable` to `INSERT` the data from the `sp1500_tbl` tibble into the `security` table.

```{r load_security_data}
dbAppendTable(secdb, "security", sp1500_tbl)
```

And, we can check with the following raw `sql` chunks to verify that the security data are loaded correctly. I also take the opportunity here to display a `SQL` join between the `security` table and the `gics` table to display the sector name for the security which is accessed by using the first two digits of the `sub_industry_code` for the security. Note that `SQL` does all of the appropriate number and character and back to number conversion that is required to make the `WHERE` clause join logic work correctly.

```{sql security_select}
SELECT uid,
       DATE(start_date) AS start_date,
       DATE(end_date) AS end_date,
       symbol,
       security.name AS name,
       gics.name AS sector
FROM   security,
       gics
WHERE  gics.level = 'SCT'
  AND  gics.code = SUBSTR(security.sub_industry_code,1,2);
```

```{r db_disconnect, include = FALSE}
dbDisconnect(secdb)
```
